---
title: "Data Analysis SMARTTESTER®"
author: "Hannah Rausch (adapted from Ivan Calandra & Lisa Schunk)"
date: "`r format(Sys.time(), usetz = TRUE)`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cerulean
    highlight: pygments
    number_sections: true
  github_document: 
    toc: true
    toc_depth: 3
    html_preview: false
bibliography: grateful-refs.bib
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all", knit_root_dir = rprojroot::find_rstudio_root_file()) })
---

```{r Knitr Options, include = FALSE}
knitr::opts_chunk$set(comment = NA, message = FALSE, indent = "", error = TRUE)
```

---

# Goal of the script
Data analysis of SMARTTESTER sensor data


```{r}
dir_in <- "a_raw_data_smarttester"
dir_out <- "d_output_data_smarttester"
```

Raw data must be located in `r paste0("~/", dir_in)`.  
Formatted data will be saved in `r paste0("~/", dir_out)`.
The knit directory for this script is the project directory.

---

# Load packages
```{r}
pack_to_load <- c("tidyverse", "R.utils", "openxlsx", "tools", "patchwork", "doBy", "ggrepel", "ggplot2", "grateful", "viridis")
sapply(pack_to_load, library, character.only = TRUE, logical.return = TRUE) 
```

---

# Get name, path and information of the file
```{r}
data_file <- list.files(dir_in, pattern = "\\.xlsx$", full.names = TRUE)
data_file
```


# Load data into R object
```{r}
imp_data <- read.xlsx(data_file)
str(imp_data)
```

The imported file is: "`r paste0("~/", data_file)`"  

---

# Plot each of the selected numeric variable 
## Plots showing the strokes as lines 
```{r}
# split the data in the individual 4 samples
sp <- split(imp_data, imp_data[["Sample"]])
str(sp)

# select parameters to plot
y_var <- c("Force", "Friction", "Depth", "Velocity")

# Plot every 40th stroke
for (i in seq_along(sp)) {
  # create a sequence of every 40th stroke, always including the first and the last strokes
   seq_st <- seq(1, length(unique(sp[[i]][["Stroke"]])), by = 40) %>% 
    c(max(unique(sp[[i]][["Stroke"]])))
  
  #
  dat_i_40th <- filter(sp[[i]], Stroke %in% seq_st)
  dat_i_50 <- filter(sp[[i]], Stroke %in% 1:50)
  
  # Pivot to longer format for facet plots
  data_long_40th <- select(dat_i_40th, all_of(c("Relative_pos", y_var, "Stroke"))) %>%
             pivot_longer(all_of(y_var), names_to = "parameter", values_to = "value")
  str(data_long_40th)
  data_long_50 <- select(dat_i_50, all_of(c("Relative_pos", y_var, "Stroke"))) %>%
             pivot_longer(all_of(y_var), names_to = "parameter", values_to = "value") 
  str(data_long_50)
  
  # Axes labels with units
  param.labs <- c("Force [N]", "Friction [N]", "Depth [mm]", "Velocity [mm/s]")
  names(param.labs) <- y_var
    
  # Plot
  p1 <- ggplot(data = data_long_40th, aes(x = Relative_pos, y = value, color = Stroke)) +
    geom_line(aes(group = Stroke), alpha = 0.3) + 
    # Wrap around parameters with free y-scales
    facet_wrap(~ parameter, scales = "free_y", ncol = 1, labeller = labeller(parameter = param.labs)) +
    
    labs(x = "Distance [mm]", y = NULL) +
    # reverse the legend starting with 0 going to 2000 strokes    
    scale_color_viridis(trans = "reverse") + 
    # change the 'Step-number' in the x-legend  
    scale_x_continuous(breaks=c(0, 35, 70, 105, 140)) +
    theme_classic()
  
  # plot only the first 50 strokes per sample  
  p2 <- ggplot(data = data_long_50, aes(x = Relative_pos, y = value, color = Stroke)) +
    geom_line(aes(group = Stroke), alpha = 0.3) + 
    # Wrap around parameters with free y-scales
    facet_wrap(~ parameter, scales = "free_y", ncol = 1, labeller = labeller(parameter = param.labs)) +
    
    labs(x = "Distance [mm]", y = NULL) +
    # reverse the legend starting with 0 going to 2000 strokes    
    scale_color_viridis(trans = "reverse") + 
    # change the 'Step-number' in the x-legend  
    scale_x_continuous(breaks=c(0, 35, 70, 105, 140)) +
    theme_classic()

  # patchwork plot
  p <- p2 + p1 + plot_annotation(title = names(sp)[i]) 
  print(p)
  # save to PDF
  file_out <- paste0(file_path_sans_ext(basename(data_file)), "_plot_", names(sp)[i], ".pdf") %>%
    gsub("b_", "d_", .)
  ggsave(filename = file_out, plot = p, path = dir_out, device = "pdf")
} 
```

---

# Descriptive statistics
## Create function to compute the statistics at once
```{r Function}
NMinMaxMeanMedSd <- function(x){
  y <- x[!is.na(x)]     # Exclude NAs
  n_test <- length(y)   # Sample size (n)
  min_test <- min(y)    # Minimum
  max_test <- max(y)    # Maximum
  mean_test <- mean(y)  # Mean
  med_test <- median(y) # Median
  sd_test <- sd(y)      # Standard deviation
  out <- c(n_test, min_test, max_test, mean_test, med_test, sd_test) # Concatenate
  names(out) <- c("n", "min", "max", "mean", "median", "sd")         # Name values
  return(out)                                                        # Object to return
}
N.distinct <- function(x){
  y <- x[!is.na(x)]     # Exclude NAs
  n_test <- length(y)   # Sample size (n)
  unique_test <- length(unique(y)) # Number of unique values
  out <- c(n_test, unique_test) # Concatenate
  names(out) <- c("n", "distinct")         # Name values
  return(out)                                                        # Object to return
}
```

## Summary statistics
```{r}
all_stats <- summaryBy(.~ Sample, data = imp_data[c(y_var, "Sample")], FUN = NMinMaxMeanMedSd)
all_stats
stroke_stats <- summaryBy(.~ Sample, data = imp_data[c("Stroke", "Sample")], FUN = N.distinct)
stroke_stats
```

## Export statistics
```{r}
write.xlsx(list(Stroke = stroke_stats, Parameters = all_stats), 
           file = paste0(dir_out, "/d_descriptivestats_smarttester_stepmm.xlsx"))
```

---

# sessionInfo()

```{r}
sessionInfo()
```


---


# Cite R packages used

```{r, echo = FALSE}
cite_packages(pkgs = pack_to_load, output = "paragraph", include.RStudio = TRUE, out.dir = "Script")
```


## References