---
title: "Data Analysis SMARTTESTER®"
author: "Hannah Rausch (adapted from Ivan Calandra & Lisa Schunk)"
date: "`r format(Sys.time(), usetz = TRUE)`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: cerulean
    highlight: pygments
    number_sections: true
  github_document: 
    toc: true
    toc_depth: 3
    html_preview: false
bibliography: grateful-refs.bib
knit: (function(inputFile, encoding) {
  rmarkdown::render(inputFile, encoding = encoding, output_format = "all", knit_root_dir = rprojroot::find_rstudio_root_file()) })
---

```{r Knitr Options, include = FALSE}
knitr::opts_chunk$set(comment = NA, message = FALSE, indent = "", error = TRUE)
```

---

# Goal of the script
Analysis of the SMARTTESTER sensor data containing the parameters depth, force, friction and velocity.
This is achieved by creating line graphs displaying the data of each parameter for each of the four samples.
Additionally, descriptive statistics of each parameter for each sample is retrieved.
A bibliography of the R packages used is found at the end of the script.

```{r}
dir_in <- "a_raw_data_smarttester"
dir_out <- "b_output_data_smarttester"
```

Raw data must be located in `r paste0("~/", dir_in)`.  
Formatted data will be saved in `r paste0("~/", dir_out)`.
The knit directory for this script is the project directory.

---

# Load packages
```{r}
pack_to_load <- c("tidyverse", "R.utils", "openxlsx", "tools", "patchwork", "doBy", "ggplot2", "grateful", "viridis")
sapply(pack_to_load, library, character.only = TRUE, logical.return = TRUE) 
```

---

# Get name, path and information of the file
```{r}
data_file <- list.files(dir_in, pattern = "\\.xlsx$", full.names = TRUE)
data_file
```


# Load data into R object

```{r}
imp_data <- read.xlsx(data_file)
str(imp_data)
```

The imported file is: "`r paste0("~/", data_file)`"  

---

# Calculate new parameters
## Relative depth
The lower the recorded value for `Depth`, the deeper the tool has penetrated into the contact material. This scale is not intuitive on a plot, so we decided to calculate the relative depth.  

For `Sample` s, the relative depth at `Stroke` i & `Step` j is then defined as the `Depth` at `Stroke` i and `Step` j minus the maximum depth across all strokes:  
$Rel.Depth_{s,i,j} = Depth_{s,i,j} - Max.depth_s$  

```{r}
imp_data <- imp_data %>% 
            group_by(Sample) %>% 
            mutate(Rel.Depth = Depth - max(Depth, na.rm = TRUE))
```

With this calculation, the scale of the `Rel.Depth` parameter is in a more intuitive way on a plot: lower (negative) values indicate deeper penetration.  
```{r}
head(select(imp_data, Sample, Stroke, Step, Depth, Rel.Depth), n = 10L)
```

While the maximum depth should theoretically be at the beginning of the experiment (Stroke 1 & Step 1) for each sample, it is not the case. This is why we chose the maximum recorded `Depth` per sample ($Max.depth_s$ ), rather than the first `Depth` measurement, as the reference for the relative depth on each sample.  
The maximum depth for each sample has been recorded at these strokes and steps:
```{r}
imp_data %>% group_by(Sample) %>% slice_max(Depth)
```

## Relative position
The `Position` value along the X axis is recorded across each stroke at 10 Hz (i.e. 10 readings per second). Due to inaccuracies, the absolute X position at each `Step` might slightly differ from one stroke to another. Additionally, the starting position is not 0. We therefore calculate here the relative position.

For `Sample` s and `Stroke` i, the relative position at `Step` j is then defined as the `Position` at `Step` j minus the `Position` at `Step` 1:  
$Rel.Position_{s,i,j} = Position_{s,i,j} - Position_{s,i,1}$  

```{r}
imp_data <- imp_data %>% 
            group_by(Sample, Stroke) %>% 
            mutate(Rel.Position = Position - Position[1])
head(select(imp_data, Sample, Stroke, Step, Position, Rel.Position), n = 10L)
```

## Check results
```{r}
imp_data <- as.data.frame(imp_data)
head(imp_data, n = 10L)
str(imp_data)
```

---

# Plot the sensor data of each parameter with strokes as line graphs
```{r}
# Split the data into the individual 4 samples.
sp <- split(imp_data, imp_data[["Sample"]])
str(sp)

# Select parameters to plot.
y_var <- c("Force", "Friction", "Rel.Depth", "Velocity")

# Create axes labels with units.
param.labs <- c("Force [N]", "Friction [N]", "Relative depth [mm]", "Velocity [mm/s]")
names(param.labs) <- y_var

# Plot
for (i in seq_along(sp)) {
  # Create a sequence of every 40th stroke per sample that includes the first and the last strokes.
  seq_st <- seq(1, length(unique(sp[[i]][["Stroke"]])), by = 40) %>% 
            c(max(unique(sp[[i]][["Stroke"]])))
  
  # Create a subset of every 40th stroke and a subset of the first 50 strokes per sample.
  dat_i_40th <- filter(sp[[i]], Stroke %in% seq_st)
  dat_i_50 <- filter(sp[[i]], Stroke %in% 1:50)
  
  # Pivot to longer format for facet plots.
  data_long_40th <- select(dat_i_40th, all_of(c("Rel.Position", y_var, "Stroke"))) %>%
                    pivot_longer(all_of(y_var), names_to = "parameter", values_to = "value")
  str(data_long_40th)
  data_long_50 <- select(dat_i_50, all_of(c("Rel.Position", y_var, "Stroke"))) %>%
                  pivot_longer(all_of(y_var), names_to = "parameter", values_to = "value") 
  str(data_long_50)
  
  # Plot every 40th stroke.
  p1 <- ggplot(data = data_long_40th, aes(x = Rel.Position, y = value, color = Stroke)) +
        geom_line(aes(group = Stroke), alpha = 0.3) + 
        
        # Wrap around parameters with free y-scales.
        facet_wrap(~ parameter, scales = "free_y", ncol = 1, labeller = labeller(parameter = param.labs)) +
        labs(x = "Distance [mm]", y = NULL) +
        
        # Reverse the legend starting with 0 going to 2000 strokes.
        # Use the viridis color scale.
        scale_color_viridis(trans = "reverse") + 
        
        # Change the 'Distance [mm]' in the x-legend.
        scale_x_continuous(breaks = c(0, 35, 70, 105, 140)) +
        theme_minimal()
  
  # Plot only the first 50 strokes per sample.
  p2 <- ggplot(data = data_long_50, aes(x = Rel.Position, y = value, color = Stroke)) +
        geom_line(aes(group = Stroke), alpha = 0.3) + 
        facet_wrap(~ parameter, scales = "free_y", ncol = 1, labeller = labeller(parameter = param.labs)) +
        labs(x = "Distance [mm]", y = NULL) +
        scale_color_viridis(trans = "reverse") + 
        scale_x_continuous(breaks = c(0, 35, 70, 105, 140)) +
        theme_minimal()

  # Patchwork the plots.
  p <- p2 + p1 + plot_annotation(title = names(sp)[i]) 
  print(p)
  
  # Save the plot as a PDF.
  file_out <- paste0(file_path_sans_ext(basename(data_file)), "_plot_", names(sp)[i], ".pdf") %>%
              gsub("b_", "d_", .)
  ggsave(filename = file_out, plot = p, path = dir_out, device = "pdf")
} 
```


# Plot the sensor data of each parameter with strokes as line graphs
```{r}
seq_st_all <- vector(mode = "list", length = length(sp))
data_40th_list <- vector(mode = "list", length = length(sp))

for (i in seq_along(sp)) {
  seq_st_all[[i]] <- seq(1, length(unique(sp[[i]][["Stroke"]])), by = 40) %>% 
                     c(max(unique(sp[[i]][["Stroke"]])))
  
  # Create a subset of every 40th stroke and a subset of the first 50 strokes per sample.
  data_40th_list[[i]] <- filter(sp[[i]], Stroke %in% seq_st_all[[i]])
}

data_40th_all <- do.call(rbind, data_40th_list)

p3 <- ggplot(data_40th_all, aes(x = Rel.Position, y = Rel.Depth, color = Stroke)) +
      geom_line(aes(group = Stroke), alpha = 0.3) +
      facet_wrap(~ Sample) +
      labs(x = "Distance [mm]", y = NULL) +
      scale_color_viridis(trans = "reverse") + 
      scale_x_continuous(breaks = c(0, 35, 70, 105, 140)) +
      theme_minimal()
plot(p3)

# Save the plot as a PDF.
file_out <- paste0(file_path_sans_ext(basename(data_file)), "_plot_", names(sp)[i], ".pdf") %>%
            gsub("b_", "d_", .)
ggsave(filename = file_out, plot = p3, path = dir_out, device = "pdf")
```

---

# Descriptive statistics
## Create functions to compute the statistics in bulk
```{r Function}
# n, min, max, mean, median and sd
NMinMaxMeanMedSd <- function(x){
  y <- x[!is.na(x)]     # Exclude NAs
  n_test <- length(y)   # Sample size (n)
  min_test <- min(y)    # Minimum
  max_test <- max(y)    # Maximum
  mean_test <- mean(y)  # Mean
  med_test <- median(y) # Median
  sd_test <- sd(y)      # Standard deviation
  out <- c(n_test, min_test, max_test, mean_test, med_test, sd_test) # Concatenate
  names(out) <- c("n", "min", "max", "mean", "median", "sd")         # Name values
  return(out)                                                        # Object to return
}

# n and number of distinct values
N.distinct <- function(x){
  y <- x[!is.na(x)]                # Exclude NAs
  n_test <- length(y)              # Sample size (n)
  unique_test <- length(unique(y)) # Number of unique values
  out <- c(n_test, unique_test)    # Concatenate
  names(out) <- c("n", "distinct") # Name values
  return(out)                      # Object to return
}
```

## Summary statistics
```{r}
# For all parameters
all_stats <- summaryBy(.~ Sample, data = imp_data[c(y_var, "Sample")], FUN = NMinMaxMeanMedSd)
all_stats

# For stroke
stroke_stats <- summaryBy(.~ Sample, data = imp_data[c("Stroke", "Sample")], FUN = N.distinct)
stroke_stats
```

## Export statistics
Export to XLSX with two sheets
```{r}
write.xlsx(list(Stroke = stroke_stats, Parameters = all_stats), 
           file = paste0(dir_out, "/descriptivestats_smarttester_stepmm.xlsx"))
```

---

# sessionInfo()
```{r}
sessionInfo()
```


---


# Cite R packages used
```{r, echo = FALSE}
cite_packages(pkgs = pack_to_load, output = "paragraph", include.RStudio = TRUE, out.dir = "c_script")
```


## References